### 2023-01-19 doc/manuscript/translocation.qmd
Changes to intro text to de-emaphize evo rescue

-   Species recovery in the face of disease will depend on the ability of a species to move or change.

-   Rapid evolution termed "evolutionary rescue" can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021). (Not sure if Allie wants to go the ER direction)

-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018).

@@

-   Potential knowledge gap: Subsequent recoveries from disease-induced declines following observed genetic changes have rarely \[never?\] been documented in wildlife populations (Sea stars: Moritsch and Raimondi 2018?).

-   Mendelson et al. (2019)

-   Scheele et al. (2021)

-   Joseph and Knapp (2018)

-   Potential knowledge gap/opportunity/selling point: For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions. In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion has occurred. We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease. Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction. These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.

Describe expectations if frog populations are recovering due to evolutionary rescue.

>>>>

-   Species recovery in the face of severe disease (population growth rate \< 1 ) will depend on the ability of a species to move or change.

-   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al. 2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples? Restrict to species of conservation concern?).

-   Given the urgent need not to not only increase the size of remnant populations, but also reestablish extirpated populations, methods are critically needed by which the recovery of populations can be leveraged to facilitate landscape-scale recovery in the presence of ongoing disease. For example, the disease-caused decline of a wildlife species might ideally be followed by the eventual recovery of remnant populations via evolutionary rescue. These populations could then be used as sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al. 2019, Joseph and Knapp 2018).

-   For rescued populations to contribute to landscape-scale species recovery, the fitness benefits gained through selection should manifest beyond the natal environment, allowing for the long-term growth and persistence of incipient populations in novel environmental conditions.

-   In this study, we first document evidence of strong selection in host populations that are persisting with a deadly fungal pathogen, indicative that evolutionary rescue in response to previous pathogen invasion may have occurred. We then repeatedly translocate these "rescued" populations to non-natal habitats on the landscape and use longitudinal surveys, mark-recapture analysis, and population modeling to demonstrate that "rescued" populations recover and persist across a landscape of disease. Thus, natural selection combined with translocations promotes the recovery of a species on the brink of disease-induced extinction. These results may have broad implications for the many species worldwide that are threatened with extinction by the emergence of novel diseases.

<!--# I'm not sure the following text still makes sense to include. If we de-emphasize evolutionary rescue, it seems a stretch to use it as the overarching framework for the paper. The following is my best attempt at doing so. I've left the additional text on evolutionary rescue below for use elsewhere in the manuscript. -->

We hypothesize that MYL frogs are recovering due to evolutionary rescue, that the resistant animals in these populations could be used successfully in translocations to reestablish extirpated populations, and that reestablished populations would have a relatively high probability of long-term persistence. These hypotheses generate the following expectations:

<!--# Extra ER text for use in manuscript -->

-   "Evolutionary rescue" describes adaptation that occurs quickly enough to prevent extinction due to maladaptation to a new environment (Carlson et al. 2014), can reverse ongoing declines in some species (Golas et al. 2021; Searle et al. 2021).

-   ER has been proposed as a mechanism of recovery in wildlife populations decimated by disease (Epstein et al. 2016, Schiebelhut et al. 2018). This is despite the fact that when environmental change is rapid (e.g., arrival of a novel disease), the initial level of population maladaptation will be large, leaving little time for adaptive evolution to restore positive growth: Gomulkiewicz and Holt (1995), Bell and Gonzalez (2009), Lindsey et al. (2013).

@@
